On multilevel constraint decomposition methods \\ for nonlinear variational inequalities

### 2022-07-26 paper/mcd2.tex
sweep of superficial edits and planning

We construct a new proof of convergence which removes the assumption that the problem arises from optimization.  Furthermore we address finite element (FE) implementation in nonlinear cases, and we demonstrate the resulting performance.  In fact, the theory presented in \cite{Tai2003} is extended in four particular directions:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item We do not assume that the continuum VI problem arises from optimization of a scalar objective; examples are explored in Sections \ref{sec:vi} and \ref{sec:results}.  We prove convergence in $H^1$ norm at the same rate as the original method converges in energy (Section \ref{sec:convergence}). % THIS IS THE HOPE

\item We make the observation that, if implemented appropriately, multilevel ``up-smoothing'' is intrinsically more efficient than ``down-smoothing'' (Section \ref{sec:multilevel}).  This observation seems to be new; compare the comments on V(1,0) and V(1,1) cycles in \cite{GraeserKornhuber2009,Tai2003}.  A strong preference for up-smoothing is, apparently, special to multilevel CD methods and does not arise in corresponding unconstrained problems.

\item We make the multilevel CD algorithm more practical by addressing storage of intermediate quantities.  In Section \ref{sec:results} we show results from a full approximation storage (FAS; see \cite{Brandt1977}) implementation for nonlinear operators, one which avoids global (Newton) linearization; compare \cite{GraeserKornhuber2009}.

\item For problems of porous-media type the nonlinear operator $f$ is not known to be coercive, so the convergence theory in \emph{(i)} does not apply.  However, a full-cycle implementation of the multilevel CD method is demonstrated numerically.  By ``freezing' the solution-dependent coefficient, the operator is approximated by a coercive operator, nonlinear in general, for the duration of the V-cycle.  The resulting scheme is highly-effective for doubly-nonlinear diffusion operators (Section \ref{sec:results}).
\end{enumerate}

The iterates from a CD method are always admissible, and thus the operator need only be defined for admissible states; Section \ref{sec:results} includes a nontrivial example.  Admissible-iterate methods should permit direct solutions of certain VI problems, such as fluid-layer dynamics problems \cite{Bueler2021conservation,JouvetBueler2012}, for which non-admissible methods, such as semi-smooth methods \cite{BensonMunson2006}, at least require unnatural modifications of the operator formula.  The full-cycle scheme in \emph{(iv)} above is also designed for these problems, which are characterized by the solution of an auxiliary PDE on a domain determined inside the VI residual evaluation.

@@

@@

Suppose $\cV$ is a real, reflexive Banach space with norm $\|\cdot\|$ and topological dual space $\cV'$.  Denote the dual pairing of $\phi \in \cV'$ and $v\in\cV$ by $\ip{\phi}{v} = \phi(v)$, and note that $\|\phi\|_{\cV'} = \sup_{\|v\|=1} |\ip{\phi}{v}|$ defines a (Banach space) norm on $\cV'$.

Let $\cK \subset \cV$ be a nonempty closed and convex subset, the \emph{constraint set}; elements of $\cK$ are said to be \emph{admissible}.  For a continuous, but generally nonlinear, operator $f:\cK \to \cV'$ and \emph{source functional} $g\in \cV'$ we consider the following \emph{variational inequality} (VI) for the (exact) solution $u^*\in \cK$, if it exists:
\begin{equation}
\ip{f(u^*)}{v-u^*} \ge \ip{g}{v-u^*} \qquad \text{for all } v\in \cK. \label{eq:vi}
\end{equation}
Because $f$ is (generally) nonlinear, $g$ is not strictly needed for posing this VI, and by redefining $f$ we may take $g=0$, but the presence of $g$ is helpful to algorithms in Section \ref{sec:multilevel}.

VI \eqref{eq:vi} generalizes the nonlinear system of equations $f(u^*)=g$ to the constraint set $\cK$.  Informally, if we conceptualize the dual pairing as an inner product then \eqref{eq:vi} says that the angle between $f(u^*)-g$ and any arbitrary vector $v-u$ pointing from $u$ into $\cK$ is at most $90^\circ$.  That is, \eqref{eq:vi} says that $f(u^*)-g$ may not be zero but it points directly into $\cK$.  In particular, if $u^*$ is in the interior $\cK^\circ$ then \eqref{eq:vi} implies $f(u^*)=g$.

@@

Note that Tai \cite{Tai2003} uses ``coercive'' for what we wil call $2$-coercive.  It is easy to see that if $f$ is $p$-coercive then it is monotone, strictly monotone, and coercive, and thus the following result holds.

@@

When $f$ is monotone, \eqref{eq:vi} generalizes the problem of minimizing a convex function over $\cK$.  Suppose $F:\cK \to \RR$ is lower semi-continuous and (G\^ateau) differentiable with continuous derivative $F':\cK \to \cV'$.  Then $F$ is convex if and only if $F'$ is monotone \cite[Proposition I.5.5]{EkelandTemam1976}.  Furthermore, Proposition II.2.1 in \cite{EkelandTemam1976} shows that if $F$ is convex then \eqref{eq:vi} holds for $f=F'$ and $g=0$ if and only if
\begin{equation}
u^* = \operatorname{arg-min}_{v\in\cK} F(v). \label{eq:minimization}
\end{equation}
The CD methods of Tai \cite{Tai2003} address problem \eqref{eq:minimization} under the hypothesis that $F'$ is coercive.

@@

\begin{example}  \label{ex:plaplacian}  Suppose $a\in L^\infty(\Omega)$ such that $a(x)\ge a_0$ a.e.~for some constant $a_0>0$, and $p\ge 2$.  For $u,v \in \cV = W^{1,p}_0(\Omega)$ define $f:\cV \to \cV'$ by
\begin{equation}
\ip{f(u)}{v} = \int_\Omega a(x) |\grad u|^{p-2} \grad u \cdot \grad v\,dx. \label{eq:plaplacian}
\end{equation}
Now, if $x,y\in\RR^d$ then $(|x|^{p-2} x - |y|^{p-2} y)\cdot (x-y) \ge 2^{2-p} |x-y|^p$ \cite[see Appendix A and references therein]{Bueler2021conservation}.  Thus it follows from the Poincar\'e inequality that
    $$\ip{f(u) - f(v)}{u-v} \ge 2^{2-p} a_0 \|\grad u - \grad v\|_p^p \ge 2^{2-p} a_0 C \|u-v\|^p$$
for some $C>0$, and thus $f$ is $p$-coercive.  On the other hand, for $g\in\cV'$ define
    $$F(v) = \int_\Omega \frac{a(x)}{p} |\grad v|^p\,dx - \ip{g}{v}.$$
Then $F'(v) = f(v) - g$, $F$ is a convex functional (since $f$ is coercive), and, for any closed and convex $\cK\subset \cV$, VI problem \eqref{eq:vi} for is equivalent to optimization problem \eqref{eq:minimization}.\end{example}

The map in \eqref{eq:plaplacian} is coercive if $1<p<2$, but the proof is somewhat different \cite[Theorem 4.4]{Bueler2021conservation}.  In Section \ref{sec:results} we need only the $p\ge 2$ case.

Not all VI problems arise as in Example \ref{ex:plaplacian}, from optimization.  We give two such examples next, first a coercive and linear advection-diffusion problem, and then a nonlinear porous-medium-type problem; each is important in applications.  The first is preceded by a lemma.

@@

\section{Constraint decomposition, the basic algorithm} \label{sec:cd}

@@

Algorithms \ref{alg:basiccd-add} and \ref{alg:basiccd-mult} below state the basic CD method as iterations which improve a current iterate $u \in \cK$ by solving ``smaller'' VI problems over each subset $\cK_i$.  The output $w\in\cK$ should be closer to the solution $u^* \in \cK$ of \eqref{eq:vi}.

@@

The \textbf{for} loop in Algorithm \ref{alg:basiccd-add}, the additive or parallel CD method, can be computed in any order.  Note that $u-R_iu+\hat w_i$ involves removing the part of $u$ which lies in $\mathcal{K}_i$ before adding-back the improved subset solution $\hat w_i$.

The following multiplicative (successive) version orders the sets $\mathcal{K}_i$, and each subset solution updates the global iterate.  While Algorithm \ref{alg:basiccd-add} generalizes a Jacobi iteration, this one is like Gauss-Seidel.  The sum $\sum_{j>i} R_j u$ should be read as ``keep the parts of the current iterate which we have not yet improved.''

@@

Note the use of a damping parameter $0<\alpha\le 1$.  In Algorithm \ref{alg:basiccd-add} its role is obvious, as we keep part of the current iterate if $\alpha<1$, but in Algorithm \ref{alg:basiccd-mult} we only update the iterate ``so far'', namely $w_i$, using the same damping that will be applied in generating the final output $w$.

In each boxed VI above the argument of $f$ is an element of $\cK$, as the reader should confirm.  Likewise, by \eqref{eq:constraintdecomp} and \eqref{eq:constraintrestrictionsum} one may write the test element $v_i - \hat w_i \in \cV_i$ as a difference of admissible vectors, from $\cK$, namely
\begin{align*}
[u - R_i u + v_i] - [u - R_i u + \hat w_i] &= v_i - \hat w_i, \label{eq:admissibledifference} \\
\left[\sum_{j<i} w_j + v_i + \sum_{j>i} R_j u\right] - \left[\sum_{j<i} w_j + \hat w_i + \sum_{j>i} R_j u\right] &= v_i - \hat w_i,  \notag
\end{align*}
for the two Algorithms respectively.

Define
\begin{equation}
e_i = \hat w_i - R_i u \in \cV_i \label{eq:ithupdate}
\end{equation}
as the $i$th subset update in either Algorithm.  Then, as the reader may check, $\hat w = u + \sum_{i} e_i$ and $w = u + \alpha \sum_i e_i$.  The easy identity $\hat w = u^* + \sum_i \hat w_i - R_i u^*$ also holds.  We will use these identities in Section \ref{sec:convergence}.

% FOLLOWING CONTAINS ASPIRATIONS
The convergence results in Section \ref{sec:convergence} require substantial damping when applying additive Algorithm \ref{alg:basiccd-add}, namely we require $\alpha \le 1/m$ as does \cite{Tai2003}.  Algorithm \ref{alg:basiccd-mult} can be shown to converge without damping ($\alpha=1$).  In Section \ref{sec:results} we demonstrate practical convergence for larger ranges of $\alpha$ than suggested by the theory.

@@

\begin{example}  \label{ex:fnice} Suppose $f:\cV \to \cV'$ is linear and defined on all of $\cV$.  Furthermore suppose $f$ is local in the sense that a basis $\{\phi_i\}$ of $\cV_i$ exists, with $\supp \phi_i$ compactly-contained in $\Omega$, such that for any $z\in\mathcal{V}$ the value $\ip{f(\phi_i)}{z}$ can be computed by an integral over the support of $\phi_i$.  Considering only additive Algorithm \ref{alg:basiccd-add} for simplicity, by linearity the boxed VI over $\cK_i$ can be written as
\begin{equation}
\ip{f(e_i)}{v_i-\hat w_i} \ge \ip{\tilde g}{v_i-\hat w_i} \label{eq:linearlocalvi}
\end{equation}
where $\tilde g = g - f(u)$.  Note $e_i = \hat w_i - R_i u \in \cV_i \notin \cK$, in general, but VIs \eqref{eq:linearlocalvi} make sense because $f$ is defined over $\cV$.  Each problem \eqref{eq:linearlocalvi} can be solved using a stored residual $\ip{f(u)}{\cdot}$, namely as included here into the source term.  The VI \eqref{eq:vi} is then approximately solved in any incremental and efficient manner, by computations over the basis supports.
\end{example}

The assumption that $\ip{f(\phi_i)}{z}$ can be evaluated over the support of $\phi_i$ is satisfied when $f$ is a linear differential operator, $\cV$ is an FE space, and $\phi_i$ are hat functions; see Section \ref{sec:multilevel}.  Constrast non-local integral operators where $\ip{f(\phi_i)}{z}$ requires an integral over $\Omega$ even if $\phi_i$ is is a hat function with small support.

When $f$ has all the nice properties listed in Example \ref{ex:fnice} then a solver can take advantage of implementation efficiencies which are unavailable in general.  The actual solution process for VI \eqref{eq:linearlocalvi} is not the concern here; we are observing that the data of problem \eqref{eq:linearlocalvi}, and the cost of residual evaluation, are smaller than in the general case.  For more difficult problems additional ideas, illustrated in Section \ref{sec:multilevel}, are needed for practical implementation.  We will extend the basic method to nonlinear $f$ by applying the full approximation storage (FAS) idea of Brandt \cite{Brandt1977}, and the decomposed problems will only evaluate $f$ over (admissible) elements of $\mathcal{K}$.  However, extending the algorithm to non-local functionals $f$ is a topic for future research.

@@

renumbering the indices $i,j$ we may apply assumption \eqref{as:lipschitzresidual}, then the triangle inequality, and then assumption \eqref{as:lipschitzrestrictions}:
\begin{align}
E(\hat w,u^*) &\le C_2 Z \left(\sum_{i=0}^{m-1} \|\hat w_i - R_i u^*\|^2\right)^{1/2} = C_2 Z \left(\sum_{i=0}^{m-1} \|\hat w_i - R_i u + R_i u - R_i u^*\|^2\right)^{1/2} \label{eq:nextnextcore} \\
  &\le C_2 Z \left(Z + \left(\sum_{i=0}^{m-1} \|R_i u - R_i u^*\|^2\right)^{1/2}\right) \le C_2 Z \left(Z + C_1 \|u-u^*\|\right). \notag
\end{align}
Finally apply \eqref{eq:normlikebound} to give \eqref{eq:core}.
\end{proof}

@@

\section{Finite elements and multilevel constraint decomposition} \label{sec:multilevel}

@@

FIXME state essentially Algorithm 4.7 \cite{GraeserKornhuber2009} but with $\text{V}(\nu_1,\nu_2)$ cycles which works for linear; observe that up-smoothing is more efficient; state the FAS version which has $O(m)$ residual evaluation complexity on each level

@@

FIXME

@@

>>>>

The theory presented in \cite{Tai2003} is extended in three particular directions:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item We do not assume that the continuum VI problem arises from optimization of a scalar objective---examples are explored in Sections \ref{sec:vi} and \ref{sec:results}---but we prove convergence in $H^1$ norm at the same rate as the original method converges in energy (Section \ref{sec:convergence}). % THIS IS THE HOPE

\item We make finite element (FE) implementations of the multilevel CD algorithm more practical by addressing storage of intermediate quantities.  Specifically, in Section \ref{sec:results} we show results from a full approximation storage (FAS; see \cite{Brandt1977,Bruneetal2015}) implementation for nonlinear operators, one which avoids global (Newton) linearization; compare \cite{GraeserKornhuber2009}.

\item We observe that multilevel ``up-smoothing'' is intrinsically more efficient than ``down-smoothing'' (Section \ref{sec:multilevel}).  This observation seems to be new; compare the comments on V(1,0) and V(1,1) cycles in \cite{GraeserKornhuber2009,Tai2003}.  A strong preference for up-smoothing is special to multilevel CD methods and does not arise in unconstrained problems.
\end{enumerate}

For problems of porous-media type the nonlinear operator $f$ is not known to be coercive, so the convergence theory in \emph{(i)} does not apply.  However, a full-cycle implementation of the multilevel CD method is demonstrated numerically.  By ``freezing' the solution-dependent coefficient, the operator is approximated by a coercive operator, nonlinear in general, for the duration of the V-cycle.  The resulting scheme is highly-effective for doubly-nonlinear diffusion operators (Section \ref{sec:results}).

@@

The iterates from a CD method are always admissible, and thus the operator need only be defined for admissible states; Section \ref{sec:results} includes a nontrivial example.  Admissible-iterate methods should permit direct solutions of certain VI problems, such as fluid-layer dynamics problems \cite{Bueler2021conservation,JouvetBueler2012}, for which non-admissible methods, such as semi-smooth methods \cite{BensonMunson2006}, require unnatural modifications of the operator formula.  The full-cycle scheme in \emph{(ii)} above is also designed for these problems, which are characterized by the solution of an auxiliary PDE on a domain determined inside the VI residual evaluation.

@@

Suppose $\cV$ is a real, reflexive Banach space with norm $\|\cdot\|$ and topological dual space $\cV'$.  Denote the dual pairing of $\phi \in \cV'$ and $v\in\cV$ by $\ip{\phi}{v} = \phi(v)$, and note that $\|\phi\|_{\cV'} = \sup_{\|v\|=1} |\ip{\phi}{v}|$ defines a (Banach space) norm on $\cV'$.  Let $\cK \subset \cV$ be a nonempty closed and convex subset, the \emph{constraint set}; elements of $\cK$ are said to be \emph{admissible}.  For a continuous, but generally nonlinear, operator $f:\cK \to \cV'$ and \emph{source functional} $g\in \cV'$ we consider the following \emph{variational inequality} (VI) for the (exact) solution $u^*\in \cK$, if it exists:
\begin{equation}
\ip{f(u^*)}{v-u^*} \ge \ip{g}{v-u^*} \qquad \text{for all } v\in \cK. \label{eq:vi}
\end{equation}
Because $f$ is (generally) nonlinear, $g$ is not strictly needed for posing this VI,  but the presence of $g$ is helpful to algorithms in Section \ref{sec:multilevel}.  (By redefining $f$ we may take $g=0$.)

VI \eqref{eq:vi} generalizes the nonlinear system of equations $f(u^*)=g$ to the constraint set $\cK$.  Informally, if we conceptualize the dual pairing as an inner product then \eqref{eq:vi} says that the angle between $f(u^*)-g$ and any arbitrary vector $v-u$ pointing from $u$ into $\cK$ is at most $90^\circ$.  That is, \eqref{eq:vi} says that $f(u^*)-g$ may not be zero but it points directly into $\cK$.  However, if $u^*$ is in the interior $\cK^\circ$ then \eqref{eq:vi} implies $f(u^*)=g$.

@@

It is easy to see that if $f$ is $p$-coercive then it is monotone, strictly monotone, and coercive.  Note that Tai \cite{Tai2003} uses ``coercive'' for what we call $2$-coercive.  The following result holds.

@@

When $f$ is monotone, \eqref{eq:vi} generalizes the problem of minimizing a convex function over $\cK$.  In fact, suppose $F:\cK \to \RR$ is lower semi-continuous and (G\^ateau) differentiable with continuous derivative $F':\cK \to \cV'$.  Then $F$ is convex if and only if $F'$ is monotone \cite[Proposition I.5.5]{EkelandTemam1976}.  Furthermore, Proposition II.2.1 in \cite{EkelandTemam1976} shows that if $F$ is convex then \eqref{eq:vi} holds for $f=F'$ and $g=0$ if and only if
\begin{equation}
u^* = \operatorname{arg-min}_{v\in\cK} F(v). \label{eq:minimization}
\end{equation}
The CD methods of Tai \cite{Tai2003} address problem \eqref{eq:minimization} under the hypothesis that $F'$ is 2-coercive.

@@

\begin{example}  \label{ex:plaplacian}  Suppose $a\in L^\infty(\Omega)$ such that $a(x)\ge a_0$ a.e.~for some constant $a_0>0$, and $p\ge 2$.  For $u,v \in \cV = W^{1,p}_0(\Omega)$ define $f:\cV \to \cV'$ by
\begin{equation}
\ip{f(u)}{v} = \int_\Omega a(x) |\grad u|^{p-2} \grad u \cdot \grad v\,dx. \label{eq:plaplacian}
\end{equation}
Now, if $x,y\in\RR^d$ then $(|x|^{p-2} x - |y|^{p-2} y)\cdot (x-y) \ge 2^{2-p} |x-y|^p$ \cite[see Appendix A and references therein]{Bueler2021conservation}.  Thus it follows from the Poincar\'e inequality that
    $$\ip{f(u) - f(v)}{u-v} \ge 2^{2-p} a_0 \|\grad u - \grad v\|_p^p \ge 2^{2-p} a_0 C \|u-v\|^p$$
for some $C>0$, and thus $f$ is $p$-coercive.  On the other hand, for $g\in\cV'$ define
    $$F(v) = \int_\Omega \frac{a(x)}{p} |\grad v|^p\,dx - \ip{g}{v}.$$
Then $F'(v) = f(v) - g$ and $F$ is a convex functional (since $f$ is coercive).  For any closed and convex $\cK\subset \cV$, VI problem \eqref{eq:vi} for is equivalent to optimization problem \eqref{eq:minimization}.\end{example}

The map in \eqref{eq:plaplacian} is also coercive if $1<p<2$, but the proof is somewhat different \cite[Theorem 4.4]{Bueler2021conservation}.  In Section \ref{sec:results} we need only the $p\ge 2$ case.

However, not all VI problems arise from optimization.  We give two such examples next, first a coercive and linear advection-diffusion problem, and then a nonlinear porous-medium-type problem; each is important in applications.  The first is preceded by a lemma.

@@

\section{Constraint decomposition method} \label{sec:cd}

@@

Algorithms \ref{alg:basiccd-add} and \ref{alg:basiccd-mult} below state the basic CD method as functions which improve a current iterate $u \in \cK$, by solving ``smaller'' VI problems over each subset $\cK_i$, so that the output $w\in\cK$ should be closer to the solution $u^* \in \cK$ of \eqref{eq:vi}.  While Algorithm \ref{alg:basiccd-add} generalizes a Jacobi iteration, Algorithm \ref{alg:basiccd-mult} generalizes Gauss-Seidel.

@@

The \textbf{for} loop in Algorithm \ref{alg:basiccd-add}, the additive or parallel CD method, can be computed in any order.  Note that $u-R_iu+\hat w_i$ involves removing the part of $u$ which lies in $\mathcal{K}_i$ before adding-back the improved subset solution $\hat w_i$.  By contrast, the following multiplicative (successive) Algorithm \ref{alg:basiccd-mult} orders the sets $\mathcal{K}_i$, and each subset solution updates the global iterate before proceeding to the next subset.  The sum $\sum_{j>i} R_j u$ should be read as ``keep the parts of the current iterate which we have not yet improved.''

@@

Note the use of a damping parameter $0<\alpha\le 1$.  In Algorithm \ref{alg:basiccd-mult} we must update the in-progress iterate $w_i$ using the same damping that is applied to the final output $w$.

In each boxed VI above the argument of $f$ is an element of $\cK$, as the reader should confirm.  On the other hand, by \eqref{eq:constraintdecomp} and \eqref{eq:constraintrestrictionsum} one may also write the test vector $v_i - \hat w_i \in \cV_i$ as a difference of admissible vectors from $\cK$, namely, for the two Algorithms respectively:
\begin{align*}
[u - R_i u + v_i] - [u - R_i u + \hat w_i] &= v_i - \hat w_i, \label{eq:admissibledifference} \\
\left[\sum_{j<i} w_j + v_i + \sum_{j>i} R_j u\right] - \left[\sum_{j<i} w_j + \hat w_i + \sum_{j>i} R_j u\right] &= v_i - \hat w_i.  \notag
\end{align*}

Now define
\begin{equation}
e_i = \hat w_i - R_i u \in \cV_i, \label{eq:ithupdate}
\end{equation}
the $i$th subset increment in either Algorithm.  As the reader may check, $\hat w = u + \sum_{i} e_i$ and $w = u + \alpha \sum_i e_i$, and $\hat w = u^* + \sum_i \hat w_i - R_i u^*$ also holds.  We will use these identities in Section \ref{sec:convergence}.

% FOLLOWING CONTAINS ASPIRATIONS
The convergence results in Section \ref{sec:convergence} require substantial damping when applying additive Algorithm \ref{alg:basiccd-add}, namely we require $\alpha \le 1/m$ as in \cite{Tai2003}.  By constrast, Algorithm \ref{alg:basiccd-mult} can be shown to converge without damping ($\alpha\le 1$).  In Section \ref{sec:results} we demonstrate practical convergence for larger ranges of $\alpha$ than suggested by the theory.

@@

\begin{example}  \label{ex:fnice} Suppose $f:\cV \to \cV'$ is linear and defined on all of $\cV$.  Furthermore suppose $f$ is local in the sense that a basis $\{\phi_i\}$ of $\cV_i$ exists, with $\supp \phi_i$ compactly-contained in $\Omega$, such that for any $z\in\mathcal{V}$ the value $\ip{f(\phi_i)}{z}$ can be computed by an integral over the support of $\phi_i$.  Considering only additive Algorithm \ref{alg:basiccd-add} for simplicity, by linearity the boxed VI over $\cK_i$ can be written as
\begin{equation}
\ip{f(e_i)}{v_i-\hat w_i} \ge \ip{\tilde g}{v_i-\hat w_i} \label{eq:linearlocalvi}
\end{equation}
where $\tilde g = g - f(u)$.  Note $e_i = \hat w_i - R_i u \in \cV_i \notin \cK$, in general, but VIs \eqref{eq:linearlocalvi} make sense because $f$ is defined over $\cV$.  Each problem \eqref{eq:linearlocalvi} can be solved using a stored residual $\ip{f(u)}{\cdot}$, namely the quantity included here into the source term.  The VI \eqref{eq:vi} is then approximately solved in any incremental and efficient manner, by computations over the basis supports.
\end{example}

The assumption that $\ip{f(\phi_i)}{z}$ can be evaluated over the support of $\phi_i$ is satisfied when $f$ is a differential operator, $\cV$ is an FE space, and $\phi_i$ are hat functions; see Section \ref{sec:multilevel}.  Constrast non-local integral operators where $\ip{f(\phi_i)}{z}$ requires an integral over $\Omega$ even if $\phi_i$ is a hat function with small support.

When $f$ has all the nice properties listed in Example \ref{ex:fnice} then a solver can take advantage of implementation efficiencies which are unavailable in general.  The actual solution process for VI \eqref{eq:linearlocalvi} is not the concern here; we are observing that the data of problem \eqref{eq:linearlocalvi}, and the cost of residual evaluation, are smaller than in the general case.  For more difficult problems additional ideas, illustrated in Section \ref{sec:multilevel}, are needed for practical implementation.  We will extend the basic method to nonlinear $f$ by applying the full approximation storage (FAS) idea of Brandt \cite{Brandt1977}.  However, extending the algorithm to non-local functionals $f$ is a topic for future research.

@@

renumbering the indices $i,j$ we may apply assumption \eqref{as:lipschitzresidual}, then the triangle inequality, and then assumption \eqref{as:lipschitzrestrictions}:
\begin{align}
E(\hat w,u^*) &\le C_2 Z \left(\sum_{i=0}^{m-1} \|\hat w_i - R_i u^*\|^2\right)^{1/2} = C_2 Z \left(\sum_{i=0}^{m-1} \|\hat w_i - R_i u + R_i u - R_i u^*\|^2\right)^{1/2} \label{eq:nextnextcore} \\
  &\le C_2 Z \left(Z + \left(\sum_{i=0}^{m-1} \|R_i u - R_i u^*\|^2\right)^{1/2}\right) \le C_2 Z \left(Z + C_1 \|u-u^*\|\right). \notag
\end{align}
Finally apply Lemma \ref{lem:normlike} to give \eqref{eq:core}.
\end{proof}

@@

\section{Finite elements implementation} \label{sec:fe}

@@

\section{Nonlinear multilevel constraint decomposition} \label{sec:multilevel}

FIXME state the FAS version which has $O(m)$ residual evaluation complexity on each level

FIXME observe that up-smoothing is more efficient

@@

FIXME show results for 2D Firedrake implementation of NMCD for $p$-Laplacian obstacle problem (Example \ref{ex:plaplacian}), advection-diffusion obstacle problem (Example \ref{ex:advectiondiffusion}), and porous-media (Example \ref{ex:porous})

@@

\appendix
\section{Reductions of the nonlinear multilevel constraint decompostion algorithm}

FIXME by removing obstacle reduce to ordinary FAS as in \cite{Bruneetal2015}

FIXME by assuming $N$ is linear, reduce to Algorithm 4.7 in \cite{GraeserKornhuber2009} but with $\text{V}(\nu_1,\nu_2)$ cycles which works for linear
